\bf \outdated{Longitudinal Principal Manifold Estimation} \revision{Noise-corrected hippocampal shape and volume trajectories in Alzheimer's disease using longitudinal principal manifold estimation}

### 2025-02-10 manuscript/paper/main.tex
complete textual revisions

\newpage
\spacingset{1.9}
\section{Introduction}

Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) \citep{knopmanAlzheimerDisease2021}. Frequently, interest centers around longitudinal changes in one or more neurological substructures. These structural changes can be observed in magnetic resonance imaging (MRI) data \citep{crainiceanu2016tutorial}. For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with  cognitive impairment either due to AD or other causes. Image segmentation enables the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time. Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard. However, this approach is highly time and resource intensive. When analyzing data from studies with a large number of images, this approach may impose prohibitive costs. Automated segmentation approaches, including FSL-FIRST, have been introduced to address these concerns \citep{patenaudeBayesianModelShape2011}. While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.

Failures of automatic segmentation algorithms may have dramatic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study. In a study comparing the accuracy of FIRST and FreeSurfer, another automated segmentation method, using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing the scans with failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation. Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach, and may be particularly influential when using automated segmentation methods. Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data. To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.

Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} \citep{fefferman2016testing}. Manifold learning approaches have been applied in various scientific fields, e.g., single-cell biology \citep{ding2023learning} and robotics \citep{gao2023k}. In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis. Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures. Nonlinear manifold learning methods have been extensively studied in the literature. However, many popular approaches, such as Isomap \citep{tenenbaumGlobalGeometricFramework2000}, locally linear embedding \citep{roweisNonlinearDimensionalityReduction2000}, and Laplacian eigenmaps \citep{belkin2003laplacian}, are not applicable in longitudinal studies. These methods include the time dimension in the dimension reduction process, causing the time dimension to be incorporated into the low-dimensional parameterization and preventing meaningful interpretation of the time dimension.

Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal functional PCA method proposed by \cite{greven2011longitudinal}. As for nonlinear methods, \cite{wolzManifoldLearningBiomarker2010} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space. However, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously. Additionally, \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability. Finally, \cite{busch2023Multiview} developed the T-PHATE manifold learning algorithm to account for the autocorrelation present in functional MRI (fMRI) data, but the temporal modeling also depends on assumptions specific to the properties of fMRI data.

In this article, we propose a new longitudinal approach to estimating nonlinear manifolds over time. Rather than treating the time dimension the same as the space dimensions, we introduce a novel process that maintains the interpretability of the time dimension. Specifically, we use the method proposed by \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. Unlike most popular manifold learning methods, this approach results in an analytic and computationally efficient functional representation of the manifold estimate (see equation \eqref{eq: spline representation of the PME}). We build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between time points. Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, and 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.

@@

The remainder of this article is laid out as follows. Section \ref{s:PME} reviews the principal manifold framework proposed by \cite{mengPrincipalManifoldEstimation2021}. In Section \ref{s:LPME}, we adapt the principal manifold framework for longitudinal settings, and show our proposed novel algorithm for dimension reduction using the longitudinal principal manifold framework. Section \ref{s:simulations} demonstrates the performance of this approach on simulated data. In Section \ref{s:application}, we apply our proposed method to estimate smooth time dependent surfaces of the hippocampus and thalamus in ADNI participants. The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.

@@

The framework for principal manifolds originated with the concept of principal curves \citep{hastiePrincipalCurves1989}, which are essentially curves that pass through the middle of a data cloud. Motivated by the penalization approach presented by \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework, which extends principal curves to arbitrarily high intrinsic dimensions. In order to present the longitudinal PME approach, we start by introducing notation used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional spaces, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the 3-dimensional data cloud ($D=3$), while the 2-dimensional smooth surface of the region is the underlying manifold of interest ($d=2$; for example, see Figure S1 in the online supplement). For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. Given a $d$-variable function $u$, $\nabla^{\otimes 2} u$ denotes the Hessian matrix of $u$, defined as $r \to \left( \frac{\partial ^2 u}{\partial r_i \partial r_j}(r) \right)_{1 \leq i, j \leq d}$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Many manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for one- and higher-dimensional intrinsic spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are defined as follows.
\begin{definition}
  \label{def:principal_manifolds}
  Let $\mathbf{X}$ denote a random $D$-vector associated with the probability distribution $\mathbb{P}$ such that $\mathbf{X}$ has compact support, $\text{supp}(\mathbb{P})$, and finite second moments. Under some assumptions on limiting behavior and smoothness on function $f$, given a $\lambda \in [0, \infty)$, define the following functional:
  \begin{align}\label{eq:pme_kappa}
\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathbb{E}\|\mathbf{X} - f(\pi_f(\mathbf{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2.
  \end{align}
  The principal manifold of $\mathbf{X}$ with the tuning parameter $\lambda$ is the manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ if
  $f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f)$,
where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u : \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, and the projection function of $f$ satisfies $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.
\end{definition}

@@

The functional $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in equation \eqref{eq:pme_kappa} consists of two elements. The first element is the expected squared distance from data points $\mathbf{X}$ in the cloud to their projections $f(\pi_f(\mathbf{X}))$ on the manifold $M_f^d$. The second element $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$ imposes the smoothness/curvature penalty on the estimated manifold. The coefficient $\lambda$ is a tuning parameter. In this setting,  $\pi_{f_{\lambda}^{*}}(\mathbf{X})$ maps the $D$-vector $\mathbf{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space. Thus, even though manifold learning is commonly considered an unsupervised learning approach, \cite{mengPrincipalManifoldEstimation2021} show that finding the function that minimizes $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in fact takes the form of a penalized regression problem, making this a supervised learning problem. \cite{mengPrincipalManifoldEstimation2021} also show that this function takes spline form
\begin{align}\label{eq: spline representation of the PME}
    f_{(n+1), l}(r) = \sum_{j=1}^N s_{j, l} \times \eta_{4-d}\left(r - \pi_{f_{(n)}}(\mu_{j, N})\right) + \sum_{k=1}^{d + 1}\alpha_{k, l} \times p_k(r), \ \ \text{ for } \ l = 1, 2, \dots, D,
\end{align}
under the constraint $\sum_{j=1}^{N}s_{j, l} \times p_{k}\left(\pi_{f_{(n)}}(\mu_{j, N})\right) = 0$ for all $k = 1, 2, \dots, d + 1$ and $l = 1, 2, \dots, D$, where $\mu_{j, N}, j = 1, \dots, N$ denote points that summarize the data cloud. Importantly, $N$ tends to be much smaller than the number of points in the data cloud \citep[fig. 3]{mengPrincipalManifoldEstimation2021}, which indicates that the representation in equation \eqref{eq: spline representation of the PME} is computationally efficient. This spline function is specified by coefficients $s_{j, l}$ and $\alpha_{k, l}$ for $j = 1, \dots, N$, $k = 1, \dots, d+1$, and $l = 1, \dots, D$. These coefficients will be used to smooth over principal manifold estimates obtained at several time points in Section \ref{s:LPME}.

@@

Hereafter, we utilize the following notation to describe the LPME framework for generic low and high dimensions, $d$ and $D$. Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ (i.e., each $x_{it}\in\mathbb{R}^D$) for each image $i$, where $I_t$ denotes the number of observations available at each time point $t$, with $T$ total time points. As in the setting described above, at each time point $t$, these observations lie in the vicinity of a $d$-dimensional manifold and are corrupted by $D$-dimensional noise. In this setting, the manifold to estimate is represented by the following, where $t$ coincides with the time dimension
\begin{align}\label{eq: F form}
    \begin{aligned}
        F:\ \ & \mathbb{R}^d\times\mathbb{R} \rightarrow \mathbb{R}^D\times\mathbb{R},\ \ \  (t,\mathbf{r})\mapsto F(t,\mathbf{r})=:f_t(\mathbf{r}).
    \end{aligned}
\end{align}
We define a longitudinal principal manifold as follows.
\begin{definition}
  \label{def:lpme} Given a collection $\mathbf{X} = \{ \mathbf{X}_t \}_{t=1}^T$ of data clouds observed at a series of time points, where $\mathbf{X}_t$ is the random  $D$-vector observed at time $t$, and tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, we define the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ as follows
\begin{align}\label{eq:newKappa}
      \notag\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &:= \int_\mathbb{R} \mathbb{E}\left\|\mathbf{X}_t - f_t\left(\pi_{f_t}(\mathbf{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}F\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}F\right\|_{L^2(\mathbb{R}^d)}^2 \, dt,
\end{align}
\sloppy where $f_t(\mathbf{r})=F(t,\mathbf{r})$ is a continuous function of the form in equation \eqref{eq: F form}, satisfying $\lim_{\|\mathbf{r}\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(t,\mathbf{r})\|_{\mathbb{R}^{D}\times\mathbb{R}} = \infty$, and its coordinates are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u: \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$. Then, the manifold $M_{F^{*}}^{d+1}:=\{F^*(t,\mathbf{r}): (t,\mathbf{r})\in \mathbb{R}^d\times\mathbb{R}\}$ is the longitudinal principal manifold of $\mathbf{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
\end{definition}

@@

The next step in the LPME algorithm for the estimation of the longitudinal principal manifold is initialization. The PME algorithm uses Isomap \citep{tenenbaumGlobalGeometricFramework2000} to develop an initial parameterization, which the PME algorithm proceeds to iteratively improve. Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points. Here, consistency refers to the similarity of the parameterizations of similarly shaped point clouds. However, using Isomap to find parameterizations of similarly shaped manifolds (i.e., the surfaces of brain regions observed at consecutive time points) is likely to result in inconsistent parameterizations. For example, the following two drastically different expressions parameterize the same curve: $x_1 = r$, $x_2 = \sin(r + \frac{\pi}{2})$ and $x_1 = -r$, $x_2 = \sin(\frac{\pi}{2} - r)$. Because Isomap tends to yield inconsistent parameterizations when applied to each time point separately, a different approach is needed for the LPME algorithm.

@@

%\newpage
\bibliographystyle{jasa}
\bibliography{references}

>>>>

\newpage
\spacingset{1.9}
\section{Introduction}
Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) \citep{knopman2021Alzheimer}. Frequently, interest centers around longitudinal changes in one or more neurological substructures. These structural changes can be observed in magnetic resonance imaging (MRI) data \citep{crainiceanu2016tutorial}. For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with  cognitive impairment either due to AD or other causes. Image segmentation enables the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time. Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard. However, this approach is highly time and resource intensive. When analyzing data from studies with a large number of images, this approach may impose prohibitive costs. Automated segmentation approaches, including FSL-FIRST, have been introduced to address these concerns \citep{patenaude2011Bayesian}. While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.

Failures of automatic segmentation algorithms may have dramatic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study. In a study comparing the accuracy of FIRST and FreeSurfer, another automated segmentation method, using a large sample, \cite{mulder2014Hippocampal} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing the scans with failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation. \revision{Similarly, \cite{akudjedu2018comparative} found that FIRST and FreeSurfer showed strong correlation with manual segmentation for the caudate, but relatively poor absolute agreement and moderate relative consistency when applied to the hippocampus. When comparing estimated subcortical volumes obtained using FIRST and FreeSurfer in two consecutive yearly study visits, \cite{quilis-sancho2020comparative} found that considerable variability was present in the relative differences in subcortical volume estimates, even after removing outliers from analysis. Furthermore, this variability was much greater for FIRST than FreeSurfer, with standard deviations for relative volumetric difference in the left and right hippocampus reaching 12.2\% and 11.3\%, respectively \citep{quilis-sancho2020comparative}.} Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach, and may be particularly influential when using automated segmentation methods. Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data. To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.

Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} \citep{fefferman2016testing}. Manifold learning approaches have been applied in various scientific fields, e.g., single-cell biology \citep{ding2023learning} and robotics \citep{gao2023k}. In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yue2016Parameterization} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis. Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures. Nonlinear manifold learning methods have been extensively studied in the literature. However, many popular approaches, such as Isomap \citep{tenenbaum2000Global}, locally linear embedding \citep{roweis2000Nonlinear}, and Laplacian eigenmaps \citep{belkin2003laplacian}, are not applicable in longitudinal studies. These methods include the time dimension in the dimension reduction process, causing the time dimension to be incorporated into the low-dimensional parameterization and preventing meaningful interpretation of the time dimension.

Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal functional\outdated{PCA} \revision{principal component analysis} method proposed by \cite{greven2011longitudinal}. As for nonlinear methods, \cite{wolz2010Manifold} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space. However, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously. Additionally, \cite{louis2019Riemannian} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. \outdated{However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.}\revision{However, this approach assumes that the progression trajectory for each subject must be parallel to a straight reference trajectory in some latent space, which may be overly restrictive in the case of heterogeneous conditions like AD.} Finally, \cite{busch2023Multiview} developed the T-PHATE manifold learning algorithm to account for the autocorrelation present in functional MRI (fMRI) data, but the temporal modeling also depends on assumptions specific to the properties of fMRI data.

In this article, we propose a new longitudinal approach to estimating nonlinear manifolds over time. Rather than treating the time dimension the same as the space dimensions, we introduce a novel process that maintains the interpretability of the time dimension. Specifically, we use the method proposed by \cite{meng2021Principal} to model the appropriate principal manifold at each given time point. Unlike most popular manifold learning methods, this approach results in an analytic and computationally efficient functional representation of the manifold estimate (see equation \eqref{eq: spline representation of the PME}). We build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between time points. Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, and 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.

@@

The remainder of this article is laid out as follows. Section \ref{s:PME} reviews the principal manifold framework proposed by \cite{meng2021Principal}. In Section \ref{s:LPME}, we adapt the principal manifold framework for longitudinal settings, and show our proposed novel algorithm for dimension reduction using the longitudinal principal manifold framework. Section \ref{s:simulations} demonstrates the performance of this approach on simulated data. In Section \ref{s:application}, we apply our proposed method to estimate smooth time dependent surfaces of the hippocampus and thalamus in ADNI participants. The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.

@@

The framework for principal manifolds originated with the concept of principal curves \citep{hastie1989Principal}, which are essentially curves that pass through the middle of a data cloud. Motivated by the penalization approach presented by \cite{smola2001Regularized}, \cite{meng2021Principal} proposed the principal manifold estimation (PME) framework, which extends principal curves to arbitrarily high intrinsic dimensions. In order to present the longitudinal PME approach, we start by introducing notation used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional spaces, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the 3-dimensional data cloud ($D=3$), while the 2-dimensional smooth surface of the region is the underlying manifold of interest ($d=2$; for example, see Figure S1 in the online supplement). For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. Given a $d$-variable function $u$, $\nabla^{\otimes 2} u$ denotes the Hessian matrix of $u$, defined as $r \to \left( \frac{\partial ^2 u}{\partial r_i \partial r_j}(r) \right)_{1 \leq i, j \leq d}$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Many manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastie1989Principal} and \cite{meng2021Principal} for one- and higher-dimensional intrinsic spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are defined as follows.
\begin{definition}
  \label{def:principal_manifolds}
  Let $\mathbf{X}$ denote a random $D$-vector associated with the probability distribution $\mathbb{P}$ such that $\mathbf{X}$ has compact support, $\text{supp}(\mathbb{P})$, and finite second moments. Under some assumptions on limiting behavior and smoothness on function $f$, given a $\lambda \in [0, \infty)$, define the following functional:
  \begin{align}\label{eq:pme_kappa}
\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathbb{E}\|\mathbf{X} - f(\pi_f(\mathbf{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2.
  \end{align}
  The principal manifold of $\mathbf{X}$ with the tuning parameter $\lambda$ is the manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ if
  $f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f)$,
where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u : \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, and the projection function of $f$ satisfies $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.
\end{definition}

@@

The functional $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in equation \eqref{eq:pme_kappa} consists of two elements. The first element is the expected squared distance from data points $\mathbf{X}$ in the cloud to their projections $f(\pi_f(\mathbf{X}))$ on the manifold $M_f^d$. The second element $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$ imposes the smoothness/curvature penalty on the estimated manifold. The coefficient $\lambda$ is a tuning parameter. In this setting,  $\pi_{f_{\lambda}^{*}}(\mathbf{X})$ maps the $D$-vector $\mathbf{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space. Thus, even though manifold learning is commonly considered an unsupervised learning approach, \cite{meng2021Principal} show that finding the function that minimizes $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in fact takes the form of a penalized regression problem, making this a supervised learning problem. \cite{meng2021Principal} also show that this function takes spline form
\begin{align}\label{eq: spline representation of the PME}
    f_{(n+1), l}(r) = \sum_{j=1}^N s_{j, l} \times \eta_{4-d}\left(r - \pi_{f_{(n)}}(\mu_{j, N})\right) + \sum_{k=1}^{d + 1}\alpha_{k, l} \times p_k(r), \ \ \text{ for } \ l = 1, 2, \dots, D,
\end{align}
under the constraint $\sum_{j=1}^{N}s_{j, l} \times p_{k}\left(\pi_{f_{(n)}}(\mu_{j, N})\right) = 0$ for all $k = 1, 2, \dots, d + 1$ and $l = 1, 2, \dots, D$, where $\mu_{j, N}, j = 1, \dots, N$ denote points that summarize the data cloud. Importantly, $N$ tends to be much smaller than the number of points in the data cloud \citep[fig. 3]{meng2021Principal}, which indicates that the representation in equation \eqref{eq: spline representation of the PME} is computationally efficient. This spline function is specified by coefficients $s_{j, l}$ and $\alpha_{k, l}$ for $j = 1, \dots, N$, $k = 1, \dots, d+1$, and $l = 1, \dots, D$. These coefficients will be used to smooth over principal manifold estimates obtained at several time points in Section \ref{s:LPME}.

@@

Hereafter, we utilize the following notation to describe the LPME framework for generic low and high dimensions, $d$ and $D$. Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ (i.e., each $x_{it}\in\mathbb{R}^D$) for each image $i$, where $I_t$ denotes the number of observations available at each time point $t$, with $T$ total time points. As in the setting described above, at each time point $t$, these observations lie in the vicinity of a $d$-dimensional manifold and are corrupted by $D$-dimensional noise. In this setting, the manifold to estimate is represented by the following, where $t$ coincides with the time dimension

\revision{Note revision in equation from $\mathbb{R}^d \times \mathbb{R} \rightarrow \mathbb{R}^D \times \mathbb{R}$ to $\mathbb{R} \times \mathbb{R}^d \rightarrow \mathbb{R} \times \mathbb{R}^D$ to match function domain.}
\begin{align}\label{eq: F form}
    \begin{aligned}
        F:\ \ & \mathbb{R}\times\mathbb{R}^d \rightarrow \mathbb{R}\times\mathbb{R}^D,\ \ \  (t,\mathbf{r})\mapsto F(t,\mathbf{r})=:f_t(\mathbf{r}).
    \end{aligned}
\end{align}
We define a longitudinal principal manifold as follows.
\begin{definition}
  \label{def:lpme} Given a collection $\mathbf{X} = \{ \mathbf{X}_t \}_{t=1}^T$ of data clouds observed at a series of time points, where $\mathbf{X}_t$ is the random  $D$-vector observed at time $t$, and tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, we define the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ as follows
\begin{align}\label{eq:newKappa}
      \notag\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &:= \int_\mathbb{R} \mathbb{E}\left\|\mathbf{X}_t - f_t\left(\pi_{f_t}(\mathbf{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}F\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}F\right\|_{L^2(\mathbb{R}^d)}^2 \, dt,
\end{align}
\sloppy where $f_t(\mathbf{r})=F(t,\mathbf{r})$ is a continuous function of the form in equation \eqref{eq: F form}, satisfying $\lim_{\|\mathbf{r}\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(t,\mathbf{r})\|_{\mathbb{R}^{D}\times\mathbb{R}} = \infty$, and its coordinates are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u: \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$. Then, the manifold $M_{F^{*}}^{d+1}:=\{F^*(t,\mathbf{r}): (t,\mathbf{r})\in \mathbb{R}^d\times\mathbb{R}\}$ is the longitudinal principal manifold of $\mathbf{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
\end{definition}

@@

The next step in the LPME algorithm for the estimation of the longitudinal principal manifold is initialization. The PME algorithm uses Isomap \citep{tenenbaum2000Global} to develop an initial parameterization, which the PME algorithm proceeds to iteratively improve. Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points. Here, consistency refers to the similarity of the parameterizations of similarly shaped point clouds. However, using Isomap to find parameterizations of similarly shaped manifolds (i.e., the surfaces of brain regions observed at consecutive time points) is likely to result in inconsistent parameterizations. For example, the following two drastically different expressions parameterize the same curve: $x_1 = r$, $x_2 = \sin(r + \frac{\pi}{2})$ and $x_1 = -r$, $x_2 = \sin(\frac{\pi}{2} - r)$. Because Isomap tends to yield inconsistent parameterizations when applied to each time point separately, a different approach is needed for the LPME algorithm.

@@

\newpage
\bibliographystyle{jasa}
\bibliography{references}
